Lattices of acyclic pipe dreams

### 2023-03-15 acyclicFacets/draft/latticePipeDreams.tex
add figures sweeping and insertion algorithms

%Two pipe dreams are related by a \defn{flip} if they only differ by the position of a crossing and a contact (in particular, they define the same permutation).
A contact~$c$ is \defn{flippable} if the two pipes passing through contact~$c$ have a crossing~$x$.
The \defn{flip} exchanges the contact~$c$ with the crossing~$x$.
%See \cref{fig:pipeDreams}.
The flip is \defn{increasing} if the contact~$c$ is weakly southwest of the crossing~$x$.
For example, \cref{fig:pipeDreams} illustrates an increasing flip from the left pipe dream to the right pipe dream.
%
The \defn{increasing flip graph} is the graph of increasing flips on~$\pipeDreams(\omega)$.
It is clearly a directed acyclic graph, and it has a unique source and a unique sink \cite{PilaudPocchiola}, called the \defn{greedy} and \defn{antigreedy} pipe dreams, and denoted~$\greedyPipeDream$ and~$\antiGreedyPipeDream$.
\vincent{In \cite{PilaudStump-ELlabelings}, these are called positive and negative greedy facets. Also the notations are different...}
The \defn{increasing flip poset} is the reflexive and transitive closure of the increasing flip graph on~$\pipeDreams(\omega)$.

@@

Our first algorithm to construct~$\insertion{\pi}{\omega}$ is a \defn{sweeping algorithm}, inspired by the algorithm to compute greedy pipe dreams~\cite{PilaudPocchiola, PilaudStump-ELlabelings}.
An extension of this algorithm to subword complexes will be discussed in \cref{subsec:sweepingAlgorithmSubwordComplexes}, and appeared independently in~\cite{JahnStump}.

\begin{proposition}
\label{prop:sweepingAlgorithm}
For any permutations~$\pi,\omega \in \fS_n$ such that~$\pi < \omega$ in weak order, the unique pipe dream~$P \in \acyclicPipeDreams(\omega)$ such that~$\pi \in \linearExtensions(P)$ can be constructed by sweeping the triangular shape in any perturbation of the northwest direction and placing a crossing when sweeping a vertex~$v$ of the grid where pipe~$i$ arrives horizontally and pipe~$j$ arrives vertically if and only if
\begin{itemize}
\item $i < j$ and~$\omega^{-1}(i) > \omega^{-1}(j)$, and 
\item $\pi^{-1}(i) > \pi^{-1}(j)$ or vertex~$v$ lies in column~$\omega^{-1}(j)$.
\end{itemize}
See \cref{fig:sweepingAlgorithm}.
\end{proposition}

\begin{example}
\label{exm:sweepingAlgorithm}
\cref{fig:sweepingAlgorithm} illustrates the sweeping algorithm for the permutations~${\pi = 513264}$ and~${\omega = 561324}$.
The algorithm has 15 steps, but we have grouped together steps $2$ to $9$ (second arrow) and steps $12$ to $13$ (fifth arrow) as they have the same justification.
We place contacts in steps $1$, $10$, $12$, $13$ and~$15$ since~$\omega^{-1}(i) < \omega^{-1}(j)$, and in step~$14$ since~$i > j$.
We place crossings in steps $2$ to $9$ since~$i < j$, $\omega^{-1}(i) > \omega^{-1}(j)$ and we are in the column~$\omega^{-1}(j)$, and in step $11$ since~$i < j$, $\omega^{-1}(i) > \omega^{-1}(j)$ and~$\pi^{-1}(i) > \pi^{-1}(j)$.
%
\begin{figure}[b]
	\centerline{\includegraphics[scale=.9]{sweeping}}
	\caption{Sweeping algorithm for the permutations~$\pi = 513264$ and~$\omega = 561324$.} % See details in \cref{exm:sweepingAlgorithm}.}
	\label{fig:sweepingAlgorithm}
\end{figure}
\end{example}

@@

\begin{proposition}
\label{prop:insertionAlgorithm}
For any permutations~$\pi,\omega \in \fS_n$ such that~$\pi < \omega$ in weak order, the unique pipe dream~$P \in \acyclicPipeDreams(\omega)$ such that~$\pi \in \linearExtensions(P)$ can be constructed starting from the empty triangular shape and inserting the pipes~$\pi(1), \dots, \pi(n)$ one by one in the order of the permutation~$\pi$ as northwest as possible.
More precisely, at step~$t$, we insert a pipe starting at row~$\pi(t)$, ending at column~$\omega^{-1}(\pi(t))$, and whose northeast elbows are precisely covering the staircase of currently free southeast elbows in the rectangle~$[\pi(t)] \times [\omega^{-1}(\pi(t))]$.
%More precisely, at step~$t$, we consider the set~$X_t$ of free southeast elbows in the rectangle~$[\pi(t)] \times [\omega^{-1}(\pi(t))]$, and we insert a pipe starting at row~$\pi(t)$, ending at column~$\omega^{-1}(\pi(t))$, and whose northeast elbows are precisely covering the elbows of~$X_t$, and adding a new free southeast elbow in between any two consecutive elbows of~$X_t$
See \cref{fig:insertionAlgorithm}.
\end{proposition}

\begin{example}
\label{exm:insertionAlgorithm}
\cref{fig:insertionAlgorithm} illustrates the insertion algorithm for the permutations~${\pi = 513264}$ and~${\omega = 561324}$.
For instance, when inserting the (green) pipe~$4$ in the last step, the currently free southeast elbows are the first southeast elbow of the (blue) pipe~$3$ and the two southeast elbows of the (purple) pipe~$2$.
%
\begin{figure}[b]
	\centerline{\includegraphics[scale=.9]{insertion}}
	\caption{Insertion algorithm for the permutations~$\pi = 513264$ and~$\omega = 561324$.}
	\label{fig:insertionAlgorithm}
\end{figure}
\end{example}

@@

\begin{example}
\label{exm:acyclicFlipvsWeakOrder}
Consider the type~$A_2$ Coxeter system, the word~$Q = \tau_1 \tau_2 \tau_1 \tau_2 \tau_1 \tau_2$ and the longest element~$\wo = \tau_1 \tau_2 \tau_1 = \tau_2 \tau_1 \tau_2$.
The subword complex~$\subwordComplex(Q, \wo)$ has eight facets, six of which are acyclic.
All the fibers of $\equiv_{Q, \wo}$ are singletons and the lattice quotient~$[e, \wo]/\equiv_{Q, \wo}$ coincides with the weak order of type~$A_2$.
Note that the two acyclic facets~$\{1,3,4\}$ and $\{3,4,6\}$ are connected by a flip but the corresponding classes do not form a cover relation in~$[e, \wo]/\equiv_{Q, \wo}$.
%
\begin{figure}[h]
	\begin{tikzpicture}[xscale=1.5,->]
	\node (123) at (0,-2.5) {$\{1,2,3\}$};
	\node (134) at (-1,-1.5) {$\{1,3,4\}$};
	\node (126) at (0.3,-.5) {$\{1,2,6\}$};
	\node (236) at (1,-1.5) {$\{2,3,6\}$};
	\node (145) at (-1,1.5) {$\{1,4,5\}$};
	\node (156) at (-0.3,.5) {$\{1,5,6\}$};
	\node (346) at (1,1.5) {$\{3,4,6\}$};
	\node (456) at (0,2.5) {$\{4,5,6\}$};
	\draw (123)--(134);
	\draw (123)--(126);
	\draw (123)--(236);
	\draw (134)--(145);
	\draw (134)--(346);
	\draw (126)--(156);
	\draw (236)--(126);
	\draw (236)--(346);
	\draw (145)--(456);
	\draw (156)--(145);
	\draw (156)--(456);
	\draw (346)--(456);
	\end{tikzpicture}
	\qquad
	\begin{tikzpicture}[xscale=1.5,->]
	\node (123) at (0,-2.5) {$\{1,2,3\}$};
	\node (134) at (-1,-1.5) {$\{1,3,4\}$};
	\node (236) at (1,-1.5) {$\{2,3,6\}$};
	\node (145) at (-1,1.5) {$\{1,4,5\}$};
	\node (346) at (1,1.5) {$\{3,4,6\}$};
	\node (456) at (0,2.5) {$\{4,5,6\}$};
	\draw (123)--(134);
	\draw (123)--(236);
	\draw (134)--(145);
	\draw (134)--(346);
	\draw (236)--(346);
	\draw (145)--(456);
	\draw (346)--(456);
	\end{tikzpicture}
	\qquad
	\begin{tikzpicture}[xscale=1.5,->]
	\node (123) at (0,-2.5) {$\{1,2,3\}$};
	\node (134) at (-1,-1.5) {$\{1,3,4\}$};
	\node (236) at (1,-1.5) {$\{2,3,6\}$};
	\node (145) at (-1,1.5) {$\{1,4,5\}$};
	\node (346) at (1,1.5) {$\{3,4,6\}$};
	\node (456) at (0,2.5) {$\{4,5,6\}$};
	\draw (123)--(134);
	\draw (123)--(236);
	\draw (134)--(145);
	\draw (236)--(346);
	\draw (145)--(456);
	\draw (346)--(456);
	\end{tikzpicture}
	\caption{The increasing flip graph on~$\subwordComplex(Q, \wo)$ (left), its restriction to acyclic facets (middle) and the Hasse diagram of the quotient~$[e, \wo]/\equiv_{Q, \wo}$ where each class is labeled by its corresponding facet of~$\subwordComplex(Q, \wo)$ (right), for the subword complex~$\subwordComplex(Q, \wo)$ where~$Q = \tau_1 \tau_2 \tau_1 \tau_2 \tau_1 \tau_2$ and~$\wo = \tau_1 \tau_2 \tau_1 = \tau_2 \tau_1 \tau_2$. Note that the increasing flip~$\{1,3,4\} \to \{3,4,6\}$ is not a cover relation of~$[e, \wo]/\equiv_{Q, \wo}$.}
	\label{fig:acyclicFlipvsWeakOrder}
\end{figure}
\end{example}

>>>>

%Two pipe dreams are related by a \defn{flip} if they only differ by the position of a crossing and a contact (in particular, they define the same permutation).
A contact~$c$ is \defn{flippable} if the two pipes passing through contact~$c$ have a crossing~$x$.
The \defn{flip} exchanges the contact~$c$ with the crossing~$x$.
%See \cref{fig:pipeDreams}.
The flip is \defn{increasing} if the contact~$c$ is weakly southwest of the crossing~$x$.
For example, \cref{fig:pipeDreams} illustrates an increasing flip from the left pipe dream to the right pipe dream.
%
The \defn{increasing flip graph} is the graph of increasing flips on~$\pipeDreams(\omega)$.
It is clearly a directed acyclic graph, and it has a unique source and a unique sink \cite{PilaudPocchiola}, called the \defn{greedy} and \defn{antigreedy} pipe dreams, and denoted~$\greedyPipeDream$ and~$\antiGreedyPipeDream$.
The \defn{increasing flip poset} is the reflexive and transitive closure of the increasing flip graph on~$\pipeDreams(\omega)$.

@@

Our first algorithm to construct~$\insertion{\pi}{\omega}$ is a \defn{sweeping algorithm}, inspired by the algorithm to compute greedy pipe dreams~\cite{PilaudPocchiola, PilaudStump-ELlabelings}.
An extension of this algorithm to subword complexes will be discussed in \cref{subsec:sweepingAlgorithmSubwordComplexes}, and appeared independently in~\cite{JahnStump}.
We say that an order on the boxes of the triangular shape is \defn{northeast compatible} if each box~$b$ is before all boxes which are located weakly northeast of~$b$.

\begin{proposition}
\label{prop:sweepingAlgorithm}
For any permutations~$\pi,\omega \in \fS_n$ such that~$\pi < \omega$ in weak order, the unique pipe dream~$P \in \acyclicPipeDreams(\omega)$ such that~$\pi \in \linearExtensions(P)$ can be constructed by sweeping the triangular shape in any northeast compatible order and placing a crossing when sweeping a vertex~$v$ of the grid where pipe~$i$ arrives horizontally and pipe~$j$ arrives vertically if and only if
\begin{itemize}
\item $i < j$ and~$\omega^{-1}(i) > \omega^{-1}(j)$, and 
\item $\pi^{-1}(i) > \pi^{-1}(j)$ or vertex~$v$ lies in column~$\omega^{-1}(j)$.
\end{itemize}
See \cref{fig:sweepingAlgorithm1,fig:sweepingAlgorithm2}.
\end{proposition}

\begin{example}
\label{exm:sweepingAlgorithm}
\cref{fig:sweepingAlgorithm1,fig:sweepingAlgorithm2} illustrate the sweeping algorithm for the exiting permutation ${\omega = 561324}$ and the input permutations~${\pi = 513264}$ and~${\pi = 512364}$ respectively.
The algorithm has 15 steps, but we have grouped together steps $2$ to $9$ (second arrow) and steps $12$ to $13$ (fifth arrow) as they have the same justification.
For~${\pi = 513264}$ in \cref{fig:sweepingAlgorithm1}, we place 
\begin{itemize}
\item contacts at steps $1$, $10$, $12$, $13$ and~$15$ since~$\omega^{-1}(i) < \omega^{-1}(j)$, and at step~$14$ since~$i > j$,
\item crossings at steps $2$ to $9$ since~$i < j$, $\omega^{-1}(i) > \omega^{-1}(j)$ and we are in the column~$\omega^{-1}(j)$, and at step $11$ since~$i < j$, $\omega^{-1}(i) > \omega^{-1}(j)$ and~$\pi^{-1}(i) > \pi^{-1}(j)$.
\end{itemize}
For~${\pi = 512364}$ in \cref{fig:sweepingAlgorithm2}, we make the same choices, except that we place 
\begin{itemize}
\item a contact at step~$11$ since~$\pi^{-1}(i) < \pi^{-1}(j)$ and we are not in column~$\omega^{-1}(j)$, 
\item a crossing at step~$14$ since~$i < j$, $\omega^{-1}(i) > \omega^{-1}(j)$ and we are in the column~$\omega^{-1}(j)$.
\end{itemize}
%
\begin{figure}[b]
	\centerline{\includegraphics[scale=.9]{sweeping1}}
	\caption{Sweeping algorithm for the permutations~$\pi = 513264$ and~$\omega = 561324$.} % See details in \cref{exm:sweepingAlgorithm}.}
	\label{fig:sweepingAlgorithm1}
\end{figure}
%
\begin{figure}[b]
	\centerline{\includegraphics[scale=.9]{sweeping2}}
	\caption{Sweeping algorithm for the permutations~$\pi = 512364$ and~$\omega = 561324$.} % See details in \cref{exm:sweepingAlgorithm}.}
	\label{fig:sweepingAlgorithm2}
\end{figure}
\end{example}

@@

\begin{proposition}
\label{prop:insertionAlgorithm}
For any permutations~$\pi,\omega \in \fS_n$ such that~$\pi < \omega$ in weak order, the unique pipe dream~$P \in \acyclicPipeDreams(\omega)$ such that~$\pi \in \linearExtensions(P)$ can be constructed starting from the empty triangular shape and inserting the pipes~$\pi(1), \dots, \pi(n)$ one by one in the order of the permutation~$\pi$ as northwest as possible.
More precisely, at step~$t$, we insert a pipe starting at row~$\pi(t)$, ending at column~$\omega^{-1}(\pi(t))$, and whose northeast elbows are precisely covering the staircase of currently free southeast elbows in the rectangle~$[\pi(t)] \times [\omega^{-1}(\pi(t))]$.
%More precisely, at step~$t$, we consider the set~$X_t$ of free southeast elbows in the rectangle~$[\pi(t)] \times [\omega^{-1}(\pi(t))]$, and we insert a pipe starting at row~$\pi(t)$, ending at column~$\omega^{-1}(\pi(t))$, and whose northeast elbows are precisely covering the elbows of~$X_t$, and adding a new free southeast elbow in between any two consecutive elbows of~$X_t$
See \cref{fig:insertionAlgorithm1,fig:insertionAlgorithm2}.
\end{proposition}

\begin{example}
\label{exm:insertionAlgorithm}
\cref{fig:insertionAlgorithm1,fig:insertionAlgorithm2} illustrate the insertion algorithm for the exiting permutation ${\omega = 561324}$ and the input permutations~${\pi = 513264}$ and~${\pi = 512364}$ respectively.
For instance, when inserting the (green) pipe~$4$ in the last step, the currently free southeast elbows are the first southeast elbow of the (blue) pipe~$3$ and the two southeast elbows of the (purple) pipe~$2$.
%
\begin{figure}[b]
	\centerline{\includegraphics[scale=.9]{insertion1}}
	\caption{Insertion algorithm for the permutations~$\pi = 513264$ and~$\omega = 561324$.}
	\label{fig:insertionAlgorithm1}
\end{figure}
%
\begin{figure}[b]
	\centerline{\includegraphics[scale=.9]{insertion2}}
	\caption{Insertion algorithm for the permutations~$\pi = 512364$ and~$\omega = 561324$.}
	\label{fig:insertionAlgorithm2}
\end{figure}
\end{example}

@@

\begin{example}
\label{exm:acyclicFlipvsWeakOrder}
Consider the type~$A_2$ Coxeter system, the word~$Q = \tau_1 \tau_2 \tau_1 \tau_2 \tau_1 \tau_2$ and the longest element~$\wo = \tau_1 \tau_2 \tau_1 = \tau_2 \tau_1 \tau_2$.
The subword complex~$\subwordComplex(Q, \wo)$ has eight facets, six of which are acyclic.
All the fibers of $\equiv_{Q, \wo}$ are singletons and the lattice quotient~$[e, \wo]/\equiv_{Q, \wo}$ coincides with the weak order of type~$A_2$.
Note that the two acyclic facets~$\{1,3,4\}$ and $\{3,4,6\}$ are connected by a flip but the corresponding classes do not form a cover relation in~$[e, \wo]/\equiv_{Q, \wo}$.
%
\begin{figure}[h]
	\begin{tikzpicture}[xscale=1.5,->]
	\node (123) at (0,-2.5) {$\{1,2,3\}$};
	\node (134) at (-1,-1.5) {$\{1,3,4\}$};
	\node (126) at (0.3,-.5) {$\{1,2,6\}$};
	\node (236) at (1,-1.5) {$\{2,3,6\}$};
	\node (145) at (-1,1.5) {$\{1,4,5\}$};
	\node (156) at (-0.3,.5) {$\{1,5,6\}$};
	\node (346) at (1,1.5) {$\{3,4,6\}$};
	\node (456) at (0,2.5) {$\{4,5,6\}$};
	\draw (123)--(134);
	\draw (123)--(126);
	\draw (123)--(236);
	\draw (134)--(145);
	\draw (134)--(346);
	\draw (126)--(156);
	\draw (236)--(126);
	\draw (236)--(346);
	\draw (145)--(456);
	\draw (156)--(145);
	\draw (156)--(456);
	\draw (346)--(456);
	\end{tikzpicture}
	\qquad
	\begin{tikzpicture}[xscale=1.5,->]
	\node (123) at (0,-2.5) {$\{1,2,3\}$};
	\node (134) at (-1,-1.5) {$\{1,3,4\}$};
	\node (236) at (1,-1.5) {$\{2,3,6\}$};
	\node (145) at (-1,1.5) {$\{1,4,5\}$};
	\node (346) at (1,1.5) {$\{3,4,6\}$};
	\node (456) at (0,2.5) {$\{4,5,6\}$};
	\draw (123)--(134);
	\draw (123)--(236);
	\draw (134)--(145);
	\draw (134)--(346);
	\draw (236)--(346);
	\draw (145)--(456);
	\draw (346)--(456);
	\end{tikzpicture}
	\qquad
	\begin{tikzpicture}[xscale=1.5,->]
	\node (123) at (0,-2.5) {$\{1,2,3\}$};
	\node (134) at (-1,-1.5) {$\{1,3,4\}$};
	\node (236) at (1,-1.5) {$\{2,3,6\}$};
	\node (145) at (-1,1.5) {$\{1,4,5\}$};
	\node (346) at (1,1.5) {$\{3,4,6\}$};
	\node (456) at (0,2.5) {$\{4,5,6\}$};
	\draw (123)--(134);
	\draw (123)--(236);
	\draw (134)--(145);
	\draw (236)--(346);
	\draw (145)--(456);
	\draw (346)--(456);
	\end{tikzpicture}
	\caption{The increasing flip graph on~$\subwordComplex(Q, \wo)$ (left), its restriction to acyclic facets (middle) and the Hasse diagram of the quotient~$[e, \wo]/\equiv_{Q, \wo}$ where each class is labeled by its corresponding facet of~$\subwordComplex(Q, \wo)$ (right), for the subword complex~$\subwordComplex(Q, \wo)$ where~$Q = \tau_1 \tau_2 \tau_1 \tau_2 \tau_1 \tau_2$ and~$\wo = \tau_1 \tau_2 \tau_1 = \tau_2 \tau_1 \tau_2$. Note that the increasing flip~$\{1,3,4\} \to \{3,4,6\}$ is not a cover relation of~$[e, \wo]/\equiv_{Q, \wo}$.}
	\label{fig:acyclicFlipvsWeakOrder}
\end{figure}
\vincent{Do we want to redo this picture using sorting networks?}
\end{example}
